---
title: "Media and Marketing Research Resources"
subtitle: "A collection of resources related to media and marketing research, curated by [Collin K. Berke](https://www.collinberke.com/)."
execute:
  include: false
  message: false
  echo: false
format:
  html:
    toc: false
    date-modified: last-modified
---

```{r}
#| label: setup
library(readr)
library(stringr)
library(dplyr)
library(glue)
library(reactable)
```

```{r}
#| label: get-data-resources
resources <- read_csv("https://raw.githubusercontent.com/collinberke/data-media-marketing-resources/refs/heads/main/resources.csv") |>
  mutate(
    LinkText = str_remove_all(Link, "https://|/$"),
    Link = glue(
      "<a href='{Link}' target='_blank'>{LinkText}</a>"
    ),
    Tags = str_replace_all(Tags, ";", ", ")
  ) |>
  select(-LinkText)
```

This resource list was last updated on `r Sys.Date()`.
Want to add a resource? File an issue or edit the `resources.csv` file on [GitHub](https://github.com/collinberke/data-media-marketing-resources) and submit a Pull Request.

```{r}
#| include: true
resources |>
  arrange(Category, Name) |>
  reactable(
    columns = list(
      Category = colDef(
        name = "Category",
        maxWidth = 180
      ),
      Name = colDef(
        name = "Name"
      ),
      Link = colDef(
        html = TRUE
      ),
      Tags = colDef(
        maxWidth = 180
      )
    ),
    searchable = TRUE,
    striped = TRUE,
    pagination = FALSE,
    filterable = TRUE,
    sortable = TRUE,
    showSortable = TRUE,
    elementId = "tbl-resources"
  )
```

### Note

This list is provided for convenience and informational purposes only.
Some resources may be open-source and freely available.
Others may require a monetary payment, subscription, or account to access.
Inclusion on this list does not represent an endorsement or guarantee of the accuracy, reliability, or content of any linked site.
Content within these links may be invalid or inaccurate.
We take no responsibility of what is contained within these links or what results from their review or use.

This list will be updated periodically.
Resources will be added and removed during these updates.

### Inspiration

This list and its structure was inspired by [Nicola Rennie's](https://nrennie.rbind.io/) data visualization and data science resources list.

Check them out:

* [Data Visualization Resources](https://nrennie.rbind.io/data-viz-resources/)
* [Data Science Resources](https://nrennie.rbind.io/data-science-resources/)
