---
title: "Notes: Validating system dependencies for Rapp"
description: ""
format:
  html:
    link-external-newwindow: true
execute:
  eval: false
date: "2026-07-04"
author: "Collin K. Berke, Ph.D."
toc: true
code-link: true
code-copy: true
from: markdown+emoji
categories:
  - notes
  - rapp
---

Recently, I've been using [Rapp](https://github.com/r-lib/Rapp) to develop CLI tools.
My intent is to [wrap these tools up within an R package](https://github.com/r-lib/Rapp#installing-launchers) for others and my future self to use.
However, some of these utilities make calls to other CLI tools installed from other sources and are not natively available via a system's terminal.
When this happens, headaches ensue, and the library needs to be installed before the utility can be used.
Thus, some validation checking is in order.
The steps of this validation checking then include:

1. Print version information (if any is available).
2. Parse the version info to check for presence of the system library.
3. If system library isn't available, print info on how to obtain it.

To achieve this, the approach I came up with includes the use of some Base R, [stringr](https://stringr.tidyverse.org/), and [cli](https://cli.r-lib.org/index.html).
Specifically, the Base R `system2()` function is used to run the system command via R.
`stringr` is used to identify if the specific package dependency is available.
`cli` is used to print useful information to the user in cases the system dependency isn't available.

Let's look at the entire function and highlight what each piece is doing:

```{r}
is_bq_installed <- function() {
  gcloud_version_info <- system2("gcloud", args = "version", stdout = TRUE)

  if (!any(stringr::str_detect(gcloud_version_info, "bq"))) {
    cli::cli_alert_danger("Google Cloud SDK not available.")
    cli::cli_alert_info(
      "See install docs at {.url https://docs.cloud.google.com/sdk/docs/install-sdk}."
    )
  }
}
```

The first step here is to print the version information, if any is available.
In my case, many of the utilities I've been working on wrap [`bq` CLI tool](https://docs.cloud.google.com/bigquery/docs/quickstarts/load-data-bq) commands.
This line makes the system call in R:

```{r}
gcloud_version_info <- system2("gcloud", args = "version", stdout = TRUE)
```

One thing that tripped me up about using `system2()` is the `args` argument is what takes any flags for the system command, unlike the older `system()` Base R function.
The other important bit here is `stdout = TRUE`.
Setting this allows for capturing of the command's output within a R character vector, which is useful in the case for later parsing.

Since I'm interested in making sure users have the dependency available, I use this set of code to do so:

```{r}
!any(stringr::str_detect(gcloud_version_info, "bq"))
```

The reason for `str_detect()` here should be pretty obvious.
However, the `!any()` may not.
``
I hope this saves someone thinking about these issues a little bit of time.
This feels like a reasonable approach, but I'm sure there's other--maybe even better ways--to achieve the same result.
If you know of any, send me a message with your ideas.





