---
title: "Notes: Validating system dependencies for Rapp"
description: "How to make sure users have the needed tools when 'shelling out'"
format:
  html:
    link-external-newwindow: true
execute:
  eval: false
date: "2026-07-09"
author: "Collin K. Berke, Ph.D."
toc: true
code-link: true
code-copy: true
from: markdown+emoji
categories:
  - notes
  - rapp
---

# Background

Recently, I've been using [Rapp](https://github.com/r-lib/Rapp) to develop CLI tools.
Now that I'm creating some useful tools (useful to me anyways), my intent is to make them accessible to other folks.
To do this, I want to [make them available within an R package](https://github.com/r-lib/Rapp#installing-launchers).
Some of these tools, however, ["shell out"](https://ropensci.org/blog/2021/09/13/system-calls-r-package/) to other CLI tools not natively installed on a user's system.
When this happens, errors are thrown, confusion sets in, headaches ensue, and the user is left blindly guessing at missing setup steps.
What if we could catch this issue right away?
What if when the user doesn't have the needed dependency they receive some useful information on how to perform the setup?
Validation checking is the answer.

::: {.callout-note}
This [post](https://ropensci.org/blog/2021/09/13/system-calls-r-package/) from `rOpenSci` warns how wrappers to CLI tools is not always a great approach.
I heed these warnings related to a "shell out".
But for simple utilities, I'm not going to [write C or C++ code](https://blog.davisvaughan.com/posts/2019-03-02-now-you-c-me/) to perform some straightforward actions a CLI tool has already implemented.
:::

Here's my thought process for this validation check:

1. Print version information for the wrapped CLI tool (if any is available).
2. Parse the version info to check for presence of the tool.
3. If the tool isn't available on the user's system, print info on how to obtain it.

# The approach

To achieve this, the approach I came up with includes the use of some Base R, the [stringr](https://stringr.tidyverse.org/), and [cli](https://cli.r-lib.org/index.html) packages.
Specifically, the Base R `system2()` function is used to run the system command via R.
`stringr` is used to identify if the specific package dependency is available via string pattern detection.
`cli` is used to print useful information to the user in cases the system dependency isn't available.

Let's break down the entire function:

```{r}
#| eval: false
is_bq_installed <- function() {
  gcloud_version_info <- system2("gcloud", args = "version", stdout = TRUE)

  if (!any(stringr::str_detect(gcloud_version_info, "bq"))) {
    cli::cli_alert_danger("Google Cloud SDK not available.")
    cli::cli_alert_info(
      "See install docs at {.url https://docs.cloud.google.com/sdk/docs/install-sdk}."
    )
  }
}
```

The first step here is to print the version information, if any is available.
In my case, many of the utilities I've been working on wrap commands from [Google Cloud's `bq` CLI tool](https://docs.cloud.google.com/bigquery/docs/quickstarts/load-data-bq).
This line makes the system call in R to get the version information:

```{r}
#| eval: false
gcloud_version_info <- system2("gcloud", args = "version", stdout = TRUE)
```

One thing that tripped me up about using `system2()` is the `args` argument is what passes the flags for the system command, unlike the older `system()` Base R function.
The other important bit here is `stdout = TRUE`.
Setting this allows the command's output to be returned to R as a character vector.
This is useful for later parsing.

Since I'm interested in making sure users have the required dependency, I use this set of code to check if it's available:

```{r}
#| eval: false
!any(stringr::str_detect(gcloud_version_info, "bq"))
```

The reason for `str_detect()` here should be pretty obvious.
It's detecting the presence of a string pattern in the character vector, 'bq' in this case.

The `any()` function returns a single `TRUE` or `FALSE` boolean if *any* of the values in the character vector are TRUE.
The use of `!` flips it.
Thus, if the "bq" string is not detected, the end result is `FALSE`, which then kicks off the control statement:

```{r}
#| eval: false
if (!any(stringr::str_detect(gcloud_version_info, "bq"))) {
  cli::cli_alert_danger("Google Cloud SDK not available.")
  cli::cli_alert_info(
    "See install docs at {.url https://docs.cloud.google.com/sdk/docs/install-sdk}."
  )
}
```

At this point, it's using some functions from the [`cli`](https://cli.r-lib.org/) package to push installation information to the user.

# Wrap up

This feels like a reasonable approach to system dependency checking when doing some simple CLI tool wrapping.
Indeed, if you're attempting to do something more advanced, I highly suggest reviewing [rOpenSci's blog post](https://ropensci.org/blog/2021/09/13/system-calls-r-package/).
I hope this saves some time for anyone facing this same issue.

# Let's connect

If you found this content useful, please share.
If you find these topics interesting and want to discuss further, let's connect:

{{< include ../../../.includes/_connect.qmd >}}
