---
title: "Notes: Checksums, they're useful"
description: "How some simple values helped speed up file comparisons while using R and Google Cloud Storage"
format:
  html:
    link-external-newwindow: true
execute:
  eval: false
date: "2026-08-06"
author: "Collin K. Berke, Ph.D."
toc: true
code-link: true
code-copy: true
from: markdown+emoji
categories:
  - notes
  - cloud computing
  - google cloud storage
  - rstats
---

In [Issue 033](https://www.collinberke.com/updates/posts/2026-07-13-hex-update-033/#what-i-found-interesting-this-week) of my weekly updates, I highlighted learning about the usefulness of [checksums](https://en.wikipedia.org/wiki/Checksum) while working with objects (e.g., data files) stored in [Google Cloud Storage](https://cloud.google.com/storage) (GCS).
Knowing how to create checksums has been quite useful, especially when working with file-based workflows that interface with cloud storage.
So, this post expands on this topic a little further.

Here's a little setup to get started:

```{r}
#| label: setup
#| eval: true
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(digest)
library(base64enc)
library(googleCloudStorageR)

post_path <- "blog/posts/2026-08-06-notes-google-cloud-storage-hash-values/"
```

I will admit, some concepts were a little out of my depth when first learning about this.
As such, I include links to resources that provide greater depth in areas I didn't fully understand.

To start, I learned what checksums are.
[Checksums](https://en.wikipedia.org/wiki/Checksum) are a type data derived from another type of digital data, like a file.
It's like a function that takes some input data, applies some type of algorithm to that data, and returns a value.
The input data, as long as it never changes, will always result in the same output value.
This property of checksums makes them useful for verifying data integrity.
This [video](https://www.youtube.com/watch?v=zD9HNrnN4VY) does a really good job explaining checksums using the R language.

# Checksums and Google Cloud Storage

Every object stored in GCS provides two hash values: crc32c_hash and the md5_hash.
These values are metadata stored along with each object, and they can be viewed by running the following command from the [Google Cloud SDK](https://cloud.google.com/sdk):

```bash
glcoud storage objects describe gs://bucket/object-name
```

If you use the [googleCloudStorageR](https://code.markedmondson.me/googleCloudStorageR/) package, you can also return these values for any object stored in a bucket using the `gcs_list_objects()` function and setting the `detail` argument to `full`.

```{r}
#| label: list-gcs-objects
googleCloudStorageR::gcs_list_objects(
  bucket = "bucket-name",
  prefix = "prefix-to-object",
  detail = "full"
)
```

Take note, though, what gets returned is not the actual hash values.
GCS converts and stores these as [Base64 encoded values](https://www.youtube.com/watch?v=aUdKd0IFl34).
More on that in a sec.

## A little more about the crc32c and md5 hash values

Hashing and hashing algorithms have their roots in cryptography, which, if you spend any amount of time researching, you'll find yourself going down a deep, deep rabbit hole quite quickly.
So, I'm going to keep this brief and very high level, but I provide links for an additional jumping off point.
The idea is the same, input data is fed into a hashing algorithm (in our case a crc32c or md5 algorithm) and an output value is returned.
The math to convert these values is all that is changing here.
[It's just a mathematical formula performed on data](https://www.youtube.com/watch?v=QZY3IjFBtFY).

Both the crc32c and md5 were created for different purposes.
The [crc32c hashing algorithm](https://github.com/google/crc32c) is useful in detecting accidental data corruption when data is sent over a network.
It's also very fast to calculate, as modern processors have [built in instructions](https://en.wikipedia.org/wiki/SSE4) to support this type of calculation.
The [MD5 hashing algorithm](https://www.ietf.org/rfc/rfc1321.txt) is a 128-bit string, which is useful in cases of verifying file integrity and to detect any corruption to a file.
The MD5 calculation is software-based, so it's not as efficient as the crc32c hash.

Many resources mentioned both values are 'crypotgraphically broken'.
That is, they no longer are useful for security purposes.
But hey, security discussions are outside the scope of this post.
So, if you're intending to use them for those purposes, there is no advice for you here.

# Calculating hash values in R

Now that we know a little more about these hash values, let's figure out how we can create them using R.
Although methods exist to do this in Base R, I found the [`digest`](https://eddelbuettel.github.io/digest/) package useful.
Here's an example using an `mtcars.csv` file.

```{r}
#| include: false
write_csv(
  mtcars,
  here(str_c(post_path, "mtcars.csv"))
)
```

```{r}
#| label: example-mtcars-digest
#| eval: true
# crc32c hash value
digest(
  here(str_c(post_path, "mtcars.csv")),
  algo = "crc32c",
  file = TRUE
)

#md5 hash value
digest(
  here(str_c(post_path, "mtcars.csv")),
  algo = "md5",
  file = TRUE
)
```

Pretty simple, just use the `digest()` function, tell it what object you would like to create a hash for, specify what algorithm you would like to apply, and then indicate you're seeking to convert an actual file (i.e., `file = TRUE`).

It's important to note here that `digest()` can create hash values for any R object using a variety of available algorithms.

# Using R to base64 encode hash values

As I mentioned, Google Cloud Storage converts these hash values into base64 encoded values.
So, to use these to make comparisons between what's on your local files vs what's stored in the cloud, we need to perform a transformation.
AI, specifically [Google Gemini](https://en.wikipedia.org/wiki/Google_Gemini) was helpful for figuring this out and allowing me to better understand the problem and how to solve it.
Without it, I wouldn't have known where to start.

I'll start with converting the crc32c hash value.
Then, I'll do the md5 hash value, which is the one I inevitably end up using the most often.

We'll also need to use the [`base64enc`](https://www.rforge.net/base64enc/) R package here.

## A little detour, bits and bytes

Before we get started, for my understanding, I had to establish a foundational understanding of some key terms.
These include:

* [Bits](https://en.wikipedia.org/wiki/Bit): the most basic unit of information in computing.
These data are generally represented as '1's and '0's.
* [Bytes](https://en.wikipedia.org/wiki/Byte): a unit of digital information that most commonly consists of eight bits.
So, for a computer, the number eight, when using an eight-bit byte, would be represented using the bits 00001000.
* [Hexadecimal](https://en.wikipedia.org/wiki/Hexadecimal) is a positional number system for representing a numeric value as base 16.

These terms are helpful in understanding how these transformations work, even though R will do the heavy lifting for us.
Still hanging on, I always found [this scene](https://www.youtube.com/watch?v=0xkP_FQUsuM&t=3s) from [The Martian](https://en.wikipedia.org/wiki/The_Martian_(film)) helpful.

I promise, this will all be important here in a second.

## crc32c to base64

So, we start with the original hash value in hex for the file.

```{r}
#| label: wrngl-crc-hex-value
#| eval: true
crc_hex_value <- digest(
  here(str_c(post_path, "mtcars.csv")),
  algo = "crc32c",
  file = TRUE
)

crc_hex_value
```

What gets outputted here is an 8-character hex string.
Now, we need to convert this hex string into four raw bytes.
The following code does this for us.

```{r}
#| label: wrngl-into-raw-bytes
#| eval: true
raw_bytes <- writeBin(strtoi(crc_hex_value, base = 16L), raw(), endian = "big")

raw_bytes
```

Then, base64 encode the value.

```{r}
#| label: wrngl-base64-problem-value
#| eval: true
base64encode(raw_bytes)
```

But, a problem exists.
In fact, if you scale this up to additional files, you might notice the same checksum value gets returned `gAAAAA==`.
Why?

This is where I learned [R still utilizes 32 bit integers](https://www.r-bloggers.com/2015/06/r-in-a-64-bit-world/).
Thus, any hex value exceeding 2,147,483,647 results in integer overflow.
`strtoi()` will fail silently in this case, as it will always return an `NA` value when this occurs.
As such, the raw bits will still be produced, but they will be produced based on an `NA` value.
This is why you'll get a repeated `gAAAAA==` value for some files.

```{r}
#| label: example-of-na-integer-overflow
#| eval: true
strtoi(crc_hex_value, base = 16L)
```

So due to this limitation, a workaround needs to be applied.
What we can do is split the hex string `eb844693` into 2-character hex pairs (`eb`, `84`, `46`, `93`).
These can then be converted into raw bytes, followed by base64 encoding of the value.
The code to do this looks like this:

```{r}
#| label: wrngl-hex-pairs-to-raw-bytes
#| eval: true
hex_pairs <- substring(crc_hex_value, seq(1, 7, 2), seq(2, 8, 2))

hex_pairs

raw_bytes <- as.raw(strtoi(hex_pairs, base = 16L))

raw_bytes
```

```{r}
#| label: wrngl-raw-bytes-to-base64
#| eval: true
gcs_crc_hash <- base64encode(raw_bytes)
```

You can then compare this value with the crc32c_hash stored in Google Cloud Storage.
I'll cover this here shortly.

If you're interested and want to see the raw stream of bits, the `rawToBits()` function can be used.

```{r}
#| label: wrngl-see-raw-bits
#| eval: true
rawToBits(raw_bytes)
```

## md5 to base64

Given that integer overflow is a potential issue with the crc32c hash value, the MD5 hash is another option.
Now it's just a matter of handling and transforming a 32-character rather than an 8-character hex string.
Once again, it all starts with the `digest` package.
This time we specify we want to use the MD5 algorithm in the `algo` argument.

```{r}
#| label: wrngl-md5-hex-string
#| eval: true
md5_hex_value <- digest(
  here(str_c(post_path, "mtcars.csv")),
  algo = "md5",
  file = TRUE
)

md5_hex_value
```

Given the size of the underlying bits, a matrix will be useful for organizational purposes.

```{r}
#| label: wrngl-md5-hex-matrix
#| eval: true
md5_hex_pairs <- matrix(
  c(substring(md5_hex_value, seq(1, 31, 2), seq(2, 32, 2))),
  ncol = 1
)

md5_hex_pairs
```

We then convert to an integer followed by a raw value.

```{r}
#| label: wrngl-md5-raw-bytes
#| eval: true
md5_raw_bytes <- as.raw(strtoi(md5_hex_pairs, base = 16L))

md5_raw_bytes
```

Lastly, we perform the `base64` encoding on the raw bytes value to match with how the value is represented in GCS.

```{r}
#| label: wrngl-md5-base64-encoded
#| eval: true
gcs_md5_values <- base64encode(md5_raw_bytes)

gcs_md5_values
```

Again, if you're interested in seeing the raw bits, use the `rawToBits` function on `md5_raw_bytes`.

```{r}
#| label: wrngl-md5-raw-bits
#| eval: true
rawToBits(md5_raw_bytes)
```

# Comparing files locally to those stored in Google Cloud Storage

Now that we have the transformation figured out, we use these values to compare files in GCS to those stored on our local file system.

To provide a more concrete example, I'm going to upload three versions of `mtcars` to a Google Cloud Storage bucket.
These three versions will be `mtcars` just split by the `cyl` variable (i.e., cylinders).
But first, we need to split the data into three separate files.
So, here's the code to do this:

```{r}
#| label: write-data-mtcars-by-cyl
group_split(mtcars, cyl) |>
  walk2(
    list(
      str_c(post_path, "cyl_4_mtcars.csv"),
      str_c(post_path, "cyl_6_mtcars.csv"),
      str_c(post_path, "cyl_8_mtcars.csv")
    ),
    \(group, file_name) write_csv(group, file_name)
  )
```

Let's now upload these files to our storage bucket.

```{r}
#| label: upload-data-to-gcs
files <- list.files(
  post_path,
  pattern = "cyl_\\d_mtcars.csv",
  full.names = TRUE
)

walk(
  files,
  \(x) gcs_upload(x, name = str_remove(x, str_c(post_path, "/")))
)
```

Then, let's get the `md5Hash` and `crc32c` values using `googleCloudStorageR`'s `gcs_list_objects()` function.
The `detail` argument set to `full` is what will return these values.

```{r}
#| label: list-data-in-gcs
gcs_list_objects(detail = "full")
```

Here's the md5hash and crc32c values for each file from Google Cloud storage:

| File | md5Hash | crc32c |
| :--- | :--- | :--- |
| `cyl_4_mtcars.csv` | `Pna3yg3PXv8U5h7tE5KHkg==` | `D1wf+g==` |
| `cyl_6_mtcars.csv` | `SATlfm2lSOMYn90Wa+gvHQ==` | `ujPTEQ==` |
| `cyl_8_mtcars.csv` | `QZaPlkfRhyf4E8T3x7d0GQ==` | `u0PTxQ==` |

To make things a little easier, here are two functions to handle the transformations for us:

```{r}
#| label: func-transform-gcs-crc
#| eval: true
transform_gcs_crc <- function(file) {
  hex_pairs <- digest::digest(
    file,
    algo = "crc32c",
    file = TRUE
  ) |>
    substring(seq(1, 7, 2), seq(2, 8, 2))

  raw_bytes <- as.raw(strtoi(hex_pairs, base = 16L))

  gcs_crc <- base64enc::base64encode(raw_bytes)

  return(gcs_crc)
}
```

```{r}
#| label: func-transform-gcs-md5
#| eval: true
transform_gcs_md5 <- function(file) {
  md5_hex_value <- digest(
    file,
    algo = "md5",
    file = TRUE
  )
  md5_hex_pairs <- matrix(
    c(substring(md5_hex_value, seq(1, 31, 2), seq(2, 32, 2))),
    ncol = 1
  )
  md5_raw_bytes <- as.raw(strtoi(md5_hex_pairs, base = 16L))

  gcs_md5 <- base64encode(md5_raw_bytes)

  return(gcs_md5)
}
```

Let's give it a try and compare.

```{r}
#| label: test-file-crc-values
#| eval: true
transform_gcs_crc(here(post_path, "cyl_4_mtcars.csv"))
transform_gcs_crc(here(post_path, "cyl_6_mtcars.csv"))
transform_gcs_crc(here(post_path, "cyl_8_mtcars.csv"))
```

```{r}
#| label: test-file-md5-values
#| eval: true
transform_gcs_md5(here(post_path, "cyl_4_mtcars.csv"))
transform_gcs_md5(here(post_path, "cyl_6_mtcars.csv"))
transform_gcs_md5(here(post_path, "cyl_8_mtcars.csv"))
```

Kablam, nailed it!

# Use cases

Now that we've walked through this transformation, some may ask why go through all this trouble?
Why even use a checksum?
Why not just compare other factors of the file like modified or created date?
It's important to remember these checksums are calculated using data within the file.
Thus, any changes to the data will impact the checksum.
Modified and created date can be the same between files, but the data within could be different.
In addition, checksums align well with several use cases.

One use case is when you're working with a file-based workflow and files lack any IDs or metadata information useful in identifying individual files.
This is especially useful when these files are being transformed for storage into a data warehouse, where you're just appending new data.
These hash values can be stored with what's in the warehouse and used as a way to skip any files that have already been processed and stored, rather than having to re-run the transformation process again for every single file.
Indeed, you are storing redundant data, but that's one tradeoff you have to make here for this to work.

Another use case is when you want to ensure the files you're working with match with what's stored in cloud storage.
Maybe files change periodically or update from time to time, and you want some simple value to validate you're working with the correct files.
Checksums can be useful to ensure the file you're working with matches the ones stored in GCS.

# Wrap up

So, there it is.
I just showed you how to create md5 and crc32c hash values for a file using R.
Then, I demonstrated how to convert md5 and crc32c character strings into bytes, bits, and then into a base64 encoded values, just like they are represented for Google Cloud Storage bucket objects.
I also learned a little more about R while writing this post, especially when it comes to 32-bit integers.

Wasn't that fun?
I sure came away with some takeaways.
AI was certainly helpful for showing me how to solve this problem.

My hope is this helps someone somewhere in the future.
Keep those checksums in mind.
They may come in handy.

# Let's connect

If you found this content useful, please share.
If you find these topics interesting and want to discuss further, let's connect:

{{< include ../../../.includes/_connect.qmd >}}
