---
title: "Exploring R Consortium ISC Grants"
date: "2024-02-26"
author: "Collin K. Berke, Ph.D."
draft: false
description: "A contribution to the 2024-02-20 #tidytuesday social data project"
toc: true
from: markdown+emoji
categories:
  - data wrangling
  - data visualization
  - tidytuesday
  - plotly
  - Tableau
---

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
library(plotly)
library(skimr)
library(tidytext)
library(here)
library(scales)
```

# Background

I've never really contributed to [`tidytuesday`](https://github.com/rfordatascience/tidytuesday/tree/master). Recently, I've been trying to spark some inspiration, so I thought contributing to this social data project would be a good start. I used this post as an opportunity to get more comfortble using [`plotly`](https://plotly.com/r/) and [Tableau](https://www.tableau.com/) for creating data visualizations. 

```{r import-data, message=FALSE, warning=FALSE}
data_isc_grants <-
  read_csv(
    'https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-02-20/isc_grants.csv'
  )
```

# Data description

The data represents information about past projects funded by the R Consortium Infrastructure Committee (ISC) Grant Program. The purpose of these grants is to support projects contributing to the R community. Learn more about the most recent round of funding by checking out their [blog post](https://www.r-consortium.org/blog/2024/02/08/r-consortium-infrastructure-steering-committee-isc-grant-program-accepting-proposals-starting-march-1st) announcing this round of grants.

The data includes columns like: `year`, `group` (i.e., funding cycle), `title`, `funded` (i.e., funding amount), and `summary`. Before creating some data visualizations, let's do some quick exploratory analysis. 

```{r view-data-str}
glimpse(data_isc_grants)
```

```{r explore-data}
skim(data_isc_grants)
```

# What's the trend for grant funding? 

Let's take a look at the funding trend by funding cycle (i.e., fall and spring).

```{r wrngl-data-by-year-group}
#| code-fold: true
data_by_year_grp <- data_isc_grants |>
  mutate(
    group = case_when(
      group == 1 ~ "Spring",
      group == 2 ~ "Fall"
    )
  ) |>
  group_by(year, group) |>
  summarise(funded = sum(funded), .groups = "drop") |>
  arrange(group, year) |>
  pivot_wider(names_from = group, values_from = funded)
```

```{r write-data-for-tableau-1, eval=FALSE, include=FALSE}
data_by_year_grp <- data_isc_grants |>
  mutate(
    group = case_when(
      group == 1 ~ "Spring",
      group == 2 ~ "Fall"
    )
  ) |>
  group_by(year, group) |>
  summarise(funded = sum(funded), .groups = "drop") |>
  arrange(group, year)

write_csv(
  data_by_year_grp,
  here("2024-02-23-tidytuesday-attempts/data_by_year_grp.csv")
)
```

```{r vis-funding-trend}
#| code-fold: true
plot_ly(
  data_by_year_grp,
  x = ~year,
  y = ~Fall,
  name = "Fall",
  type = 'scatter',
  mode = 'lines',
  line = list(width = 5),
  text = ~ paste(
    "Funding awarded: $",
    comma(Fall),
    "<br>Year: ",
    year
  ),
  hoverinfo = "text"
) |>
  add_trace(
    y = ~Spring,
    name = "Spring",
    text = ~ paste(
      "Funding awarded: $",
      comma(Spring),
      "<br>Year: ",
      year
    ),
    hoverinfo = "text"
  ) |>
  layout(
    title = list(
      text = "<b>Funding trend for R Consortium ISC grants by funding round</b>",
      xanchor = "center",
      yanchor = "top",
      font = list(family = "arial", size = 24)
    ),
    xaxis = list(title = ""),
    yaxis = list(title = "Funding amount ($US)")
  )
```

# What words are used most often within descriptions of funded projects?

Now, let's explore the words used within descriptions most often in awarded grant applications.

```{r wrngl-text-data}
#| code-fold: true
#| message: false
data_word_fund_trend <- data_isc_grants |>
  mutate(
    summary = str_remove_all(str_to_lower(summary), "[[:punct:]]"),
    summary = str_remove_all(summary, "[0-9]"),
  ) |>
  unnest_tokens(word, summary) |>
  anti_join(get_stopwords()) |>
  group_by(year) |>
  count(word) |>
  arrange(word, year) |>
  group_by(word) |>
  mutate(
    n_cume = cumsum(n)
  )
```

```{r write-data-for-tableau-2, eval=FALSE, include=FALSE}
write_csv(
  data_word_fund_trend,
  here("2024-02-23-tidytuesday-attempts/data_by_word.csv")
)
```

```{r vis-cumulative-top-words}
#| code-fold: true
#| message: false
top_words <- data_word_fund_trend |>
  ungroup() |>
  summarise(top = quantile(n_cume, .99)) |>
  pull(top)

data_top_words <- data_word_fund_trend |>
  filter(n_cume >= top_words) |>
  distinct(word)

plot_ly(
  data = data_word_fund_trend,
  x = ~year,
  y = ~n_cume,
  mode = "lines",
  line = list(color = "#d3d3d3", width = 3),
  type = "scatter",
  mode = "lines",
  name = "",
  text = ~ paste(
    "Word: ",
    word,
    "<br>Cumulative mentions: ",
    n_cume,
    "<br>Year: ",
    year
  ),
  hoverinfo = "text"
) |>
  add_lines(
    data = data_word_fund_trend |> semi_join(data_top_words),
    x = ~year,
    y = ~n_cume,
    line = list(color = "#0C2D48", width = 3),
    type = "scatter",
    mode = "lines",
    name = ""
  ) |>
  layout(
    title = list(
      text = "<b>Aiming for RConsortium grant funding? Consider using these words</b>",
      xanchor = "center",
      yanchor = "top",
      font = list(family = "arial", size = 24)
    ),
    xaxis = list(title = ""),
    yaxis = list(title = "Cumulative mentions"),
    showlegend = FALSE
  )
```

# An attempt using Tableau

To learn more about using Tableau, I took this week's data as an opportunity to learn more. [Here's](https://public.tableau.com/app/profile/collin.berke/viz/TableauTuesday_17089936427980/2024-02-20-r-consortium-grants) what I came up with.
